Designing a high-performance boundary element library with OpenCL and Numba

### 2021-04-13 opencl.tex
work through rest of opencl section

\subsection{SIMD optimized kernels}

The above description was valid for kernels that do not take advantage of CPU SIMD operations are kernels that run directly on a GPU device. For SIMD optimization on CPUs we proceed slightly differently. The corresponding kernel in principle works as described above, but batches together the interaction on one test triangle with X trial triangles, where X is either 4, 8, or 16, depending on the number of available SIMD lanes. This strategy allows us to optimize almost all floating point operations within a kernel run for SIMD operation. If the number of trial elements is not divisible by 4, 8, or 16, then the few remaining trial elements are assembled with the standard-non vectorized kernel.

Each kernel definition is stored in two variants, one with the ending \textit{\_novec.cl} and another one with the ending \textit{\_vec.cl}. The vectorized variant is configured via preprocessor directives for the desired number of vector lanes. Having to develop two OpenCL kernel codes for each operator is a certain amount of overhead. If we write a new kernel, we first implement the non-vectorized version. Then, with the help of preprocessor directives and a number of helper functions that do the actual implementation of operations depending on whether the kernel is vectorized or not, it is usually a matter of an hour or two to convert the non-vectorized kernel into a vectorized version.

Alternatively, some CPU OpenCL runtime environments have the option to try and auto-vectorize kernels by batching together work items on SIMD lanes, similarly to what we do manually. In our experience this works well for very simple kernels but often fails for more complex OpenCL kernels. This is why we decided to implement this strategy manually.

A completely different SIMD strategy is based on batching together quadrature evaluations within a single test/triangle pair. There are two disadvantages to this approach. First, it only works well if the number of quadrature points is a multiple of the available SIMD lanes. Second, other operations such as the geometry calculations for each element then can not be SIMD optimized as these are only performed once per test/triangle pair.

\subsection{Assembling the singular part of integral operators}

The assembly of the singular part of an integral operator works a bit differently. Remember that the singular part consists of triangle parts which are adjacent to each other or identical. There are $\mathcal{O}(N)$ of such pairs. We are using fully numerical quadrature rules for these integrals that are based on subdividing the four-dimensional integration domain and using transformation techniques to remove the singularities. This gives highly accurate evaluations of these integration pairs but requires a large number (typically over $1000$) quadrature points per triangle pair. Hence, we create one workgroup for each triangle pair and inside this workgroup have a number of work-items that are evaluating the quadrature rules and then at the end sum up the result together. Depending on how two triangles are in relationship to each other different types of singular quadrature rules are needed. We solve this pre-loading all possible quadrature rules onto the device and also store for each triangle pair an index into the required quadrature rule so that the kernel function can select the correct rule to evaluate. For the singular quadrature rules we did not implement separate SIMD optimized kernels as the proposed implementation is already highly efficient and requires only a fraction of the computational time of the regular quadrature rules described above. At the end, the singular integral values are either summed into the overall result matrix, or if desired by the user, stored as separate sparse matrix.

>>>>

\subsection{SIMD optimized kernels}
When we are running on a CPU and want to take advantage of available SIMD optimizations, we need to make a few modifications to our approach. The corresponding kernel works similarly to described above, but we compute a batch of interactions between one test triangle and $X$ trial triangles, where $X$ is either 4, 8, or 16 (depending on the number of available SIMD lanes). This strategy allows us to optimize almost all floating point operations within a kernel run for SIMD operation. If the number of trial elements is not divisible by 4, 8, or 16, then the few remaining trial elements are assembled with the standard-non vectorized kernel.

Each kernel definition is stored in two variants, one with the ending \texttt{\_novec.cl} and another one with the ending \texttt{\_vec.cl}. The vectorized variant is configured via preprocessor directives for the desired number of vector lanes. Having to develop two OpenCL kernel codes for each operator creates a certain amount of overhead, but once we have implemented the non-vectorized version then, with the help of preprocessor directives and a number of helper functions that do the actual implementation of operations depending on whether the kernel is vectorized or not, it is usually only a matter of an hour or two to convert the non-vectorized kernel into a vectorized version.

Alternatively, some CPU OpenCL runtime environments can (optionally) try and auto-vectorize kernels by batching together work items on SIMD lanes, similar to what we do manually. In our experience, this works well for very simple kernels but often fails for more complex OpenCL kernels. This is why we decided to implement this strategy manually.

A completely different SIMD strategy could be taken by batching together quadrature evaluations within a single test/trial pair. There are two disadvantages to this approach: first, it only works well if the number of quadrature points is a multiple of the available SIMD lanes. Second, other operations such as the geometry calculations for each element then cannot be SIMD optimized as these are only performed once per test/trial pair.

\subsection{Assembling the singular part of integral operators}
The assembly of the singular part of an integral operator works a bit differently. Remember that the singular part consists of triangle parts which are adjacent to each other or identical (the three later cases in \cref{fig:triangles}): there are $\mathcal{O}(N)$ such pairs. We are using fully numerical quadrature rules for these integrals that are based on subdividing the four-dimensional integration domain and using transformation techniques to remove the singularities. This gives highly accurate evaluation of these integration pairs but requires a large number (typically over $1000$) quadrature points per triangle pair.

For this assembly, we create one workgroup for each triangle pair. Inside this workgroup, we have a number of work-items that evaluate the quadrature rules then sum up the results. Depending on how two triangles are related to each other, different types of singular quadrature rule are needed. We solve this by pre-loading all possible quadrature rules onto the device, and also store for each triangle pair an index pointing to the required quadrature rule so that the kernel function can select the correct rule to evaluate. For the singular quadrature rules, we did not implement separate SIMD optimized kernels as the proposed implementation is already highly efficient and requires only a fraction of the computational time of the regular quadrature rules described above. At the end, the singular integral values are either summed into the overall result matrix, or (if desired by the user) stored as separate sparse matrix.
